Rheology of Cubic Blue Phases

### 2013-01-21 trunk/papers/bprheo/bprheo.tex
Submitted version
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1874 40876441-3496-0410-a613-a2fca35b2ae2

BPs were long considered as purely of academic interest due to their very narrow 
range of stability. This view has changed since the creation of polymer-stabilised and other thermally 
stabilised BPs~\cite{Kikuchi:2002,Coles:2005}, which has opened up the 
possibility of novel applications.
During the last few years considerable progress has been achieved regarding the behaviour 
of BPs in confined geometries~\cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}, under 
external fields ~\cite{Alexander:2008,Fukuda:2009,Henrich:2010a,Castles:2010,Tiribocchi:2011a}, 
and in the presence of colloidal particles~\cite{Ravnik:2011a}.
The kinetics of BP domain growth have been recently addressed ~\cite{Henrich:2010b}. 
However, our understanding of their dynamical behaviour under flow remains
very limited. The aim of this work is to address this issue by studying,
for the first time, the response of large BP samples to a shear flow.

@@

The time evolution of the fluid density and velocity are respectively governed
by the continuity equation
$\partial_t \rho = -\partial_\alpha(\rho v_\alpha)$, and
the following Navier-Stokes equation:
\begin{eqnarray}
\partial_t v_\alpha +\rho \,v_\beta \partial_\beta v_\alpha
&=& \partial_\beta \Pi_{\alpha \beta}+ \eta\, \partial_\beta [ \partial_\alpha v_\beta +\; \partial_\beta v_\alpha].
\label{NSE}
\end{eqnarray}
This emerges from the Chapman-Enskog expansion
of the Lattice Boltzmann (LB) equations 
that we solve numerically. A further term $\eta(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu v_\mu \delta_{\alpha \beta}$ that formally appears
in this expansion is negligible under the slow flows considered here
for which the the fluid motion is almost incompressible~\cite{Denniston:2001}.
$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LB units (these are discussed below, see~\cite{Henrich:2011a,Henrich:2010b}).
The thermodynamic stress tensor reads explicitly
\begin{eqnarray}
\Pi_{\alpha \beta}&=&P_0 \delta_{\alpha\beta}
-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)\nonumber\\
&-&\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta} + Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta} \nonumber\\
&+&2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
- \partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
\label{Pi}
\end{eqnarray}
and is responsible for strong non-Newtonian flow effects.
In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{Pi} reduces to the
scalar pressure as would appear in Eq.~\ref{NSE} for a Newtonian fluid.

@@

The system of coupled partial differential equations Eqs.~\ref{op-eom}
and~\ref{NSE} is solved by means of a
hybrid scheme~\cite{Marenduzzo:2007}. 
More details on the algorithm can
be found in~\cite{Denniston:2001, Denniston:2004}.

In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
phases BPI and BPII. 
The timestep and lattice spacing in lattice Boltzmann units (LBU) can be mapped
approximately to $\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units, respectively. The LB unit of stress
is equal to about $10^8$~Pa. Further details about the conversion 
from LBU to SI units can be found in ~\cite{Henrich:2011a,Henrich:2010b}.

@@

For typical simulations reported in this work, 
we chose as initial conditions thermodynamic states that are 
well inside the equilibrium region of the individual blue phase, and far away 
from the cholesteric-isotropic transition. Thus, temperature and chirality were 
$\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ for BPII, respectively.
For these parameters the total free energy density $f$ remained always negative at all flow rates
simulated. 
Since by Eq.~\ref{FE} $f=0$ for an isotropic phase with ${\mathbf Q}\equiv 0$, this means that our sytem, which is never far from equilibrium locally even under flow, always remains in a liquid crystalline state. 
We performed also selected simulations on metastable states at higher and lower temperatures, 
and at different chiralities, but did not find any significant differences in the 
general flow behaviour from that described below.
The only quantitative difference we found was that, for thermodynamic states that are closer to the phase boundary than the one we focused on (and describe below), the critical shear
rate at wich the disclination 
network broke up into chiral nematic states was lower. This is expected as these states have on average higher free energy densities and smaller order parameters than those we focus on in what follows, and cannot therefore withstand the same external forces before breaking down.

@@

Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
the top (bottom) part of the system flowing in the positive (negative) $x$-direction and the 
velocity gradient along the $y$-direction.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ LBU.
For clarity we classify various flow regimes, two in the case of BPII and three in the 
case of BPI. 
They are BPII-1 ($\gd \lesssim 3.125\e{-4}$, corresponding to ${\it Er} \lesssim 2.67$), 
BPII-2 ($4.687\lesssim \gd\lesssim 6.25\e{-4}$; $4\lesssim {\it Er} \lesssim 5.34$), ),
BPI-1 ($\gd \lesssim 1.95\e{-5}$; ${\it Er} \lesssim 0.33$), ), BPI-2 ($3.91\e{-5}\lesssim \gd \lesssim 4.687\e{-4}$; $0.67\lesssim {\it Er} \lesssim 8$) and BPI-3 ($6.25\e{-4}\lesssim \gd\lesssim 7.813\e{-4}$; $10.67 \lesssim {\it Er} \lesssim 13.33$).
Beyond these regimes, and prior to the ultimate breakdown into an isotropic phase, 
both BPI and BPII form common configurations regardless 
of the initial state. These are a cholesteric helix with helical axis 
oriented along the direction of the velocity gradient (CHG), 
and a flow-aligned nematic state (FAN).  
As is standard \cite{Henrich:2010b,Henrich:2012b} disclination lines are represented by plotting an isosurface of the scalar order parameter $q_s$. 
Typical choices are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
(Note that $q_s$ is small but non-zero at the disclination core. The director is undefinded there because the largest and second largest eigenvalues of ${\mathbf Q}$ coincide.)

@@

\begin{figure}[htpb]
\center
\includegraphics[width=0.35\textwidth]{disc-160k_run902.png}
\includegraphics[width=0.35\textwidth]{disc-164k_run902.png}
\includegraphics[width=0.35\textwidth]{disc-165k_run902.png}
\caption{Disclination network of BPII in shear flow: 
The pictures show a typical sequence of snapshots in the steady state 
at $\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity 
gradient is oriented along the vertical direction, whereas the 
horizontal direction is the flow direction. Lees-Edwards boundary 
conditions have been imposed in such a way that network moves to the 
right in the upper half and to the left in the lower half of the picture.}
\label{bp2-med}
\end{figure}

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{dir+y-250k_run949.png}
\includegraphics[width=0.495\textwidth]{dir+y-253k_run949.png}
\includegraphics[width=0.495\textwidth]{dir+y-255k_run949.png}
\includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
\caption{Director field of BPII at high shear rate: The images show snapshots at time steps 
$t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page. The colour code gives the alignment of the director field in flow direction (x) with red fully aligned and blue out of plane. 
The configuration is a helical wave that travels along the direction of vorticity.
It is translationally invariant along the gradient and flow direction.}
\label{bp2-high}
\end{figure}

@@

BPI has body-centered cubic symmetry and, unlike BPII, the disclination lines
characterising its equilibrium structure are well separated and do not 
intersect. (The quiescent state resembles the first frame in Fig.~\ref{bp1-low}
below.)
We believe that this topological difference is responsible for most of
the differences between BPI and BPII regarding their rheological response. 
We present again key aspects in an overview before we address specific features 
in more detail.

@@

The rheological response of BPI at low shear rate, $\gd\lesssim1.95\e{-5}$
(${\it Er} \lesssim 0.33$), 
is strikingly different from that of BPII and appears to show a
yield stress. An explanation for this behaviour can be found by 
looking more closely at the average shear stress (where the
background viscosity contribution has been subtracted) and 
free energy density as a function of time, as shown in Fig. \ref{bp1-fe-yield}.
When the quiescent and equilibrated BPI network begins to flow
the shear stress increases steeply and goes through a maximum.
Shortly after it goes negative (the total shear stress 
with the background viscosity contribution remains positive) 
to become positive again later on, forming a complete cycle.
The amplitude of the excursions in these stress oscillations, and
the presence of a part of the period where the thermodynamic
contribution to the stress is negative, suggest that the
BPI network is subject to large forces. Seemingly, these eventually
cause the flow-induced collapse to an amorphous network state with
an apparent yield stress. We should underscore that this
low shear regime, might depend on our choice of boundary
conditions, which in practice are equivalent to having infinitely
distant walls to which the network is anchored (see discussion
in the Appendix).
With free boundary conditions and for very low wall velocity, 
permeative flows might instead allow for
some slip of the BPI network, as in cholesterics sheared
along their axis~\cite{Marenduzzo:2006b}.

@@

The intricacy and periodic recurrence of the flow-induced conformations 
in BPI raises the question: how robust are these results to variations in 
system size? Although a pitch length of 64 LBU provides enough resolution 
for tracking the network in flow it is not clear whether $4^3=64$ unit cells 
in total are actually sufficient to avoid finite size effects.
Unfortunately the computational costs of larger simulations are still
prohibitive, given that large strains have to
be simulated in order to reach a steady state. Therefore we opted for a 
comparison with smaller runs.

@@

At high flow rates, $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$ ($10.67\lesssim {\it Er}\lesssim 13.33$), 
we observe a break-up of the flow-induced 
disclination network of regime BPI-2. However, 
in contrast to BPII, where the configuration is a travelling helical 
wave (cf. Fig. \ref{bp2-high}), BPI breaks down into a quasi 
two-dimensional, translationally invariant state, which is shown in 
Fig. \ref{bp1-high}. 
There are distinctive regions where the director field is 
predominantly oriented along the vorticity direction and 
other regions where it is oriented rather in flow-gradient plane. 
Occasionally while moving past each other these rolls resemble 
regions of local double twist.
Interestingly the critical Ericksen number for the break-up lies in 
about the same window as that of transition between BPII-1 and BPII-2.

\begin{figure}[htpb]
\includegraphics[width=0.45\textwidth]{dir3d-z-302k_run916.png}
\caption{Director field of BPI at high shear rate: The pictures show a region 
of $L_x\times L_y \times L_Z= 32^3$ lattice sites with the flow/gradient 
direction oriented along the horizontal/vertical axis. 
The colour code gives the magnitude of the z-component of the director field. 
The data show that the structure is quasi-2D, with rolling double twist
patterns in the $xy$ plane, and an approximate translational symmetry 
along the vorticity direction.}
\label{bp1-high}
\end{figure}

@@

\begin{figure}[htpb]
\includegraphics[width=0.45\textwidth]{dir3d+y-200k_run1179.png}
\caption{Flow induced cholesteric state. The orientation of the helix is 
along the gradient direction. The director field is quasi-static during 
the flow, which occurs in ``nematic layers''. Color coding indicates the 
$x$ component of the director field (see colorbar on the right).
For clarity only selected sites are shown along the $y$-direction (along which
the structure is translationally invariant).}
\label{fig:cholflow}
\end{figure}

@@

At slightly larger Ericksen numbers $0.4\lesssim Er \lesssim8$,
BPI kinematically bypasses the large stress fluctuations and 
flows with very complex, periodically recurring flow-induced conformations.
These conformations entail regularly arrange helical disclinations which 
rotate due to the shear -- a new flow-induced state which was not previously
predicted. 
The oscillatory dynamics is in some cases intermittent, and the
recurrence period for the network conformations and stress patterns
is not given by a simple formula as for BPII.
Despite their complex appearance, the flow-induced conformations are 
topologically connected to the quiescent BPI and after switching off of the 
shear flow a defect-free blue phase reforms. It is tempting to
interpret the unsteady oscillations seen in this regime as an instance of 
deterministic rheological chaos~\cite{fielding, Cates:2002}.

Just as in the case of BPII, the BPI network breaks up at larger Ericksen numbers in
a two-step process. First, at $10\lesssim\ Er\lesssim 15$, it adopts a 
quasi-two-dimensional configuration that is translationally invariant along vorticity direction. 
Then, at $16\lesssim Er \lesssim 22$, it forms a cholestric helix along the flow gradient 
direction.  Finally, at $Er\gtrsim30$ (BPI) and $Er\gtrsim10$ (BPII) the configuration
is a flow-aligned nematic state.
Although experimental evidence to support
our results is currently not available, we hope this work will inspire such experiments, and believe it can shed some light 
on the flow properties of complex liquid-crystalline phases.

@@

% tables should appear as floats within the text
%
% Here is an example of the general form of a table:
% Fill in the caption in the braces of the \caption{} command. Put the label
% that you will use with \ref{} command in the braces of the \label{} command.
% Insert the column specifiers (l, r, c, d, etc.) in the empty braces of the
% \begin{tabular}{} command.
% The ruledtabular enviroment adds doubled rules to table and sets a
% reasonable default table settings.
% Use the table* environment to get a full-width table in two-column
% Add \usepackage{longtable} and the longtable (or longtable*}
% environment for nicely formatted long tables. Or use the the [H]
% placement option to break a long table (with less control than 
% in longtable).
% \begin{table}%[H] add [H] placement to break table across pages
% \caption{\label{}}
% \begin{ruledtabular}
% \begin{tabular}{}
% Lines of table here ending with \\
% \end{tabular}
% \end{ruledtabular}
% \end{table}

>>>>

BPs were long considered as purely of academic interest due to their very narrow 
range of stability. This view has changed since the creation of polymer-stabilised and other thermally 
stabilised BPs~\cite{Kikuchi:2002,Coles:2005}, which has opened up the 
possibility of novel applications.
During the last few years considerable progress has been achieved regarding the behaviour 
of BPs in confined geometries~\cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}, under 
external fields~\cite{Alexander:2008,Fukuda:2009,Henrich:2010a,Castles:2010,Tiribocchi:2011a}, 
and in the presence of colloidal particles~\cite{Ravnik:2011a}.
The kinetics of BP domain growth have been recently addressed~\cite{Henrich:2010b}. 
However, our understanding of their dynamical behaviour under flow remains
very limited. The aim of this work is to address this issue by studying,
for the first time, the response of large BP samples to a shear flow.

@@

The time evolution of the fluid density and velocity are respectively governed
by the continuity equation
$\partial_t \rho = -\partial_\alpha(\rho v_\alpha)$, and
the following Navier-Stokes equation:
\begin{eqnarray}
\partial_t v_\alpha +\rho \,v_\beta \partial_\beta v_\alpha
&=& \partial_\beta \Pi_{\alpha \beta}+ \eta\, \partial_\beta [ \partial_\alpha v_\beta +\; \partial_\beta v_\alpha].
\label{NSE}
\end{eqnarray}
This emerges from the Chapman-Enskog expansion
of the lattice Boltzmann (LB) equations 
that we solve numerically. A further term $\eta(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu v_\mu \delta_{\alpha \beta}$ that formally appears
in this expansion is negligible under the slow flows considered here
for which the the fluid motion is almost incompressible~\cite{Denniston:2001}.
$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LB units (these are discussed below, see~\cite{Henrich:2011a,Henrich:2010b}).
The thermodynamic stress tensor reads explicitly
\begin{eqnarray}
\Pi_{\alpha \beta}&=&P_0 \delta_{\alpha\beta}
-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)\nonumber\\
&-&\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta} + Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta} \nonumber\\
&+&2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
- \partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
\label{Pi}
\end{eqnarray}
and is responsible for strong non-Newtonian flow effects.
In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{Pi} reduces to the
scalar pressure as would appear in Eq.~\ref{NSE} for a Newtonian fluid.

@@

The system of coupled partial differential equations Eqs.~\ref{op-eom}
and~\ref{NSE} is solved by means of a
hybrid scheme~\cite{Marenduzzo:2007}, which combines different numerical schemes
and provides a way to address the multiscale nature of the problem.
More details on the algorithm and the boundary conditions can
be found in~\cite{Denniston:2001, Denniston:2004, Henrich:2012a}.

In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
phases BPI and BPII.
Typical runtimes for system size $L_x\times L_y\times L_z=128^3$ on 512 processes were in the region of 18 to 24 hours.  
The timestep and lattice spacing in lattice Boltzmann units (LBU) can be mapped
approximately to $\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units, respectively. The LB unit of stress
is equal to about $10^8$~Pa. Further details about the conversion 
from LBU to SI units can be found in ~\cite{Henrich:2011a,Henrich:2010b}.

@@

For typical simulations reported in this work, 
we chose as initial conditions thermodynamic states that are 
well inside the equilibrium region of the individual blue phase, and far away 
from the cholesteric-isotropic transition. Thus, temperature and chirality were 
$\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ for BPII, respectively.
For these parameters the total free energy density $f$ remained always negative at all flow rates
simulated. 
Since by Eq.~\ref{FE} $f=0$ for an isotropic phase with ${\mathbf Q}\equiv 0$, this means that our system, which is never far from equilibrium locally even under flow, always remains in a liquid crystalline state. 
We performed also selected simulations on metastable states at higher and lower temperatures, 
and at different chiralities, but did not find any significant differences in the 
general flow behaviour from that described below.
The only quantitative difference we found was that, for thermodynamic states that are closer to the phase boundary than the one we focused on (and describe below), the critical shear
rate at which the disclination 
network broke up into chiral nematic states was lower. This is expected as these states have on average higher free energy densities and smaller order parameters than those we focus on in what follows, and cannot therefore withstand the same external forces before breaking down.

@@

Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
the top (bottom) part of the system flowing in the positive (negative) $x$-direction and the 
velocity gradient along the $y$-direction.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ LBU.
For clarity we classify various flow regimes, two in the case of BPII and three in the 
case of BPI. 
They are BPII-1 ($\gd \lesssim 3.125\e{-4}$, corresponding to ${\it Er} \lesssim 2.67$), 
BPII-2 ($4.687\lesssim \gd\lesssim 6.25\e{-4}$; $4\lesssim {\it Er} \lesssim 5.34$), ),
BPI-1 ($\gd \lesssim 1.95\e{-5}$; ${\it Er} \lesssim 0.33$), ), BPI-2 ($3.91\e{-5}\lesssim \gd \lesssim 4.687\e{-4}$; $0.67\lesssim {\it Er} \lesssim 8$) and BPI-3 ($6.25\e{-4}\lesssim \gd\lesssim 7.813\e{-4}$; $10.67 \lesssim {\it Er} \lesssim 13.33$).
Beyond these regimes, and prior to the ultimate breakdown into an isotropic phase, 
both BPI and BPII form common configurations regardless 
of the initial state. These are a cholesteric helix with helical axis 
oriented along the direction of the velocity gradient (CHG), 
and a flow-aligned nematic state (FAN).  
As is standard \cite{Henrich:2010b,Henrich:2012b} disclination lines are represented by plotting an isosurface of the scalar order parameter $q_s$. 
Typical choices are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
(Note that $q_s$ is small but non-zero at the disclination core. The director is undefined there because the largest and second largest eigenvalues of ${\mathbf Q}$ coincide.)

@@

\begin{figure}[htpb]
\center
\includegraphics[width=0.35\textwidth]{disc-160k_run902.png}
\includegraphics[width=0.35\textwidth]{disc-164k_run902.png}
\includegraphics[width=0.35\textwidth]{disc-165k_run902.png}
\caption{Disclination network of BPII in shear flow: 
The pictures show a typical sequence of snapshots in the steady state 
at $\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity 
gradient is oriented along the vertical direction (y), whereas the 
horizontal direction (x) is the flow direction. Lees-Edwards boundary 
conditions have been imposed in such a way that the network moves to the 
right in the upper half and to the left in the lower half of the picture.}
\label{bp2-med}
\end{figure}

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{dir+y-250k_run949.png}
\includegraphics[width=0.495\textwidth]{dir+y-253k_run949.png}
\includegraphics[width=0.495\textwidth]{dir+y-255k_run949.png}
\includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
\caption{Director field of BPII at high shear rate: The images show snapshots at time steps 
$t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page. The colour code gives the alignment of the director field in the flow direction (x) with red fully aligned and blue out of plane. 
The configuration is a helical wave that travels along the direction of vorticity.
It is translationally invariant along the gradient and flow direction.}
\label{bp2-high}
\end{figure}

@@

BPI has body-centred cubic symmetry and, unlike BPII, the disclination lines
characterising its equilibrium structure are well separated and do not 
intersect. (The quiescent state resembles the first frame in Fig.~\ref{bp1-low}
below.)
We believe that this topological difference is responsible for most of
the differences between BPI and BPII regarding their rheological response. 
We present again key aspects in an overview before we address specific features 
in more detail.

@@

The rheological response of BPI at low shear rate, $\gd\lesssim1.95\e{-5}$
(${\it Er} \lesssim 0.33$), 
is strikingly different from that of BPII and appears to show a
yield stress. An explanation for this behaviour can be found by 
looking more closely at the average shear stress (where the
background viscosity contribution has been subtracted) and 
free energy density as a function of time, as shown in Fig. \ref{bp1-fe-yield}.
When the quiescent and equilibrated BPI network begins to flow
the shear stress increases steeply and goes through a maximum.
Shortly after it goes negative (the total shear stress 
with the background viscosity contribution remains positive) 
to become positive again later on, forming a complete cycle.
The amplitude of the excursions in these stress oscillations, and
the presence of a part of the period where the thermodynamic
contribution to the stress is negative, suggest that the
BPI network is subject to large forces. Seemingly, these eventually
cause the flow-induced collapse to an amorphous network state with
an apparent yield stress. We should underscore that this
low shear regime might depend on our choice of boundary
conditions which in practice are equivalent to having infinitely
distant walls to which the network is anchored (see discussion
in the Appendix).
With free boundary conditions and for very low wall velocity, 
permeative flows might instead allow for
some slip of the BPI network, as in cholesterics sheared
along their axis~\cite{Marenduzzo:2006b}.

@@

The intricacy and periodic recurrence of the flow-induced conformations 
in BPI raises the question: how robust are these results to variations in 
system size? Although a pitch length of 64 LBU provides enough resolution 
for tracking the network in flow it is not clear whether $4^3=64$ unit cells 
in total are actually sufficient to avoid finite size effects.
Unfortunately the computational costs of larger simulations are still
quite prohibitive (in the region of ${\it O}(10^5)$ core hours per run at
doubled system size), given that large strains have to
be simulated in order to reach a steady state. Therefore we opted for a 
comparison with smaller runs.

@@

At high flow rates, $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$ ($10.67\lesssim {\it Er}\lesssim 13.33$), 
we observe a break-up of the flow-induced 
disclination network of regime BPI-2. However, 
in contrast to BPII, where the configuration is a travelling helical 
wave (cf. Fig. \ref{bp2-high}), BPI breaks down into a quasi 
two-dimensional, translationally invariant state, which is shown in 
Fig. \ref{bp1-high}. 
There are distinctive regions where the director field is 
predominantly oriented along the vorticity direction and 
other regions where it is oriented rather in flow-gradient plane. 
Occasionally, while moving past each other, these rolls resemble 
regions of local double twist.
Interestingly the critical Ericksen number for the break-up lies in 
about the same window as that of transition between BPII-1 and BPII-2.

\begin{figure}[htpb]
\includegraphics[width=0.45\textwidth]{dir3d-z-302k_run916.png}
\caption{Director field of BPI at high shear rate: The pictures show a region 
of $L_x\times L_y \times L_z= 32^3$ lattice sites with the flow/gradient 
direction oriented along the horizontal/vertical axis. 
The colour code gives the magnitude of the z-component of the director field. 
The data show that the structure is quasi-2D, with rolling double twist
patterns in the $xy$ plane, and an approximate translational symmetry 
along the vorticity direction.}
\label{bp1-high}
\end{figure}

@@

\begin{figure}[htpb]
\includegraphics[width=0.45\textwidth]{dir3d+y-200k_run1179.png}
\caption{Flow induced cholesteric state. The orientation of the helix is 
along the gradient direction. The director field is quasi-static during 
the flow, which occurs in ``nematic layers''. Colour coding indicates the 
$x$ component of the director field (see colorbar on the right).
For clarity only selected sites are shown along the $y$-direction (along which
the structure is translationally invariant).}
\label{fig:cholflow}
\end{figure}

@@

At slightly larger Ericksen numbers $0.4\lesssim Er \lesssim8$,
BPI kinematically bypasses the large stress fluctuations and 
flows with very complex, periodically recurring flow-induced conformations.
These conformations entail regularly arrange helical disclinations which 
rotate due to the shear -- a new flow-induced state which was not previously
predicted. 
The oscillatory dynamics is in some cases intermittent, and the
recurrence period for the network conformations and stress patterns
is not given by a simple formula as for BPII.
Despite their complex appearance, the flow-induced conformations are 
topologically connected to the quiescent BPI and after switching off the 
shear flow a defect-free blue phase reforms. It is tempting to
interpret the unsteady oscillations seen in this regime as an instance of 
deterministic rheological chaos~\cite{fielding, Cates:2002}.

Just as in the case of BPII, the BPI network breaks up at larger Ericksen numbers in
a two-step process. First, at $10\lesssim\ Er\lesssim 15$, it adopts a 
quasi-two-dimensional configuration that is translationally invariant along vorticity direction. 
Then, at $16\lesssim Er \lesssim 22$, it forms a cholesteric helix along the flow gradient 
direction.  Finally, at $Er\gtrsim30$ (BPI) and $Er\gtrsim10$ (BPII) the configuration
is a flow-aligned nematic state.
Although experimental evidence to support
our results is currently not available, we hope this work will inspire such experiments, and believe it can shed some light 
on the flow properties of complex liquid-crystalline phases.

@@

% tables should appear as floats within the text
%
% Here is an example of the general form of a table:
% Fill in the caption in the braces of the \caption{} command. Put the label
% that you will use with \ref{} command in the braces of the \label{} command.
% Insert the column specifiers (l, r, c, d, etc.) in the empty braces of the
% \begin{tabular}{} command.
% The ruledtabular environment adds doubled rules to table and sets a
% reasonable default table settings.
% Use the table* environment to get a full-width table in two-column
% Add \usepackage{longtable} and the longtable (or longtable*}
% environment for nicely formatted long tables. Or use the the [H]
% placement option to break a long table (with less control than 
% in longtable).
% \begin{table}%[H] add [H] placement to break table across pages
% \caption{\label{}}
% \begin{ruledtabular}
% \begin{tabular}{}
% Lines of table here ending with \\
% \end{tabular}
% \end{ruledtabular}
% \end{table}
